---
title: "Assignment 04 - STAT 689"
author: "Philip Anderson, panders2@tamu.edu"
date: "February 7, 2018"
output: pdf_document
---

```{r, message=FALSE}
# Import third-party modules
library("mgcv")
```

Import our data and take a look:

```{r}
fossil <- read.csv("C:/Users/Philip/Schools/TAMU/STAT_689/homework/semiparametric-regression/misc/fossil2.csv")
names(fossil) <- tolower(names(fossil))
head(fossil)
summary(fossil)
```


## Question 1
Display a scatterplot of the data and comment on which features appear interesting.

```{r}
# non-generalizable function for this data|plot only
fossil_plotter <- function() {
plot(fossil$age, fossil$strontium.ratio, pch="*", cex=1.5
     , main="Scatterplot of Strontium Ratio and Age"
     , xlab="Age"
     , ylab="Strontium Ratio"
     )
    }

fossil_plotter()
```

The features of this data that appear most interesting are:

* The clear non-linear relationship between age and Strontium ratio.  

* Some evidence of hetereoscedacticity 

  + Strontium ratio appears to have a higher variance at lower values of age.

* The necessity for a three-part piecewise regression, with a positive line from 90-105, negative from 105-112.5, and positive again from 112.5 to 125.


# Question 2
## Question 2a
Fit the fossil data using the default version of stats::smooth.spline 

```{r}
myspline <- stats::smooth.spline(x=fossil$age, y=fossil$strontium.ratio)
myspline
```


## Question 2B
### Is the fit statistically significant?
TBD

## Question 2C
Add the fit to the scatterplot 

```{r}
fossil_plotter()
lines(myspline, lwd=3.5)
```


# Question 3
Run the mgcv fit to the data with the default number of knots (8), along with 4 and 23 using the cubic spline option.

Using a simple function to save space.
```{r}
fossil_gam <- function(knots=8) {
  gam_mod <- mgcv::gam(strontium.ratio ~ s(age, k=knots, bs="cr")
                       , data=fossil
                       )
  return(gam_mod)
}

gam8 <- fossil_gam(knots=8)
gam4 <- fossil_gam(knots=4)
gam23 <- fossil_gam(knots=23)

```

## Question 3a
Which fits are statistically significant?

I am using gam.check(.) to arrive at these values.  I am not displaying the gam.check(.) output, as it renders many plots, and the plotting functionality does not appear to be disable-able.  It is also not clear from the gam.check soure code how to just extract p-values. 

```{r}
paste0("GAM with 8 knots spline p-value: 0.46")
paste0("GAM with 4 knots spline p-value: 0.08")
paste0("GAM with 23 knots spline p-value: 0.97")

```

Only the GAM fit with 4 knots has a statistically significant slope.

## Question 3b
Plot the data points on one graph only.

```{r}
# point ordering for plots
age_ord <- order(fossil$age)

# initiate plots
fossil_plotter()
lines(fossil$age[age_ord], gam8$fitted.values[age_ord], lty=1, lwd=3)
lines(fossil$age[age_ord], gam4$fitted.values[age_ord], lty=2, lwd=3)
lines(fossil$age[age_ord], gam23$fitted.values[age_ord], lty=3, lwd=3)

legend(91, 0.7073
       , c("gam4", "gam8", "gam23")
       , lty=c(2, 1, 3)
       , lwd= rep(3,3)
       )
```


## Question 3c
 The fits from all gam models are consistent with my initial observation of non-linearity.  The smoothing parameters for all fits also break the regression into three main pieces, with an ascending piece followed by a descending piece, again followed by an ascending piece.
 
# Question 4
What are the effective degrees of freedom for each mgcv fit?

```{r}
paste0("GAM with 8 knots edf: ", sum(gam8$edf) - 1) # subtract 1 for one predictor model
paste0("GAM with 4 knots edf: ", sum(gam4$edf) - 1)
paste0("GAM with 23 knots edf: ", sum(gam23$edf) - 1)

```


# Question 5
What is lambda for each fit?

```{r}
paste0("GAM with 8 knots lambda value: " , gam8$sp)
paste0("GAM with 4 knots lambda value: " , gam4$sp)
paste0("GAM with 23 knots lambda value: " , gam23$sp)

```


# Question 6
Are any of the p-values less than 0.10?

With a p-value of 0.08, only the GAM with 4 knots has an inadequate fit.  This is likely because the data displays a number of small trend fluctuations that only 4 knots are unable to capture.  We can see in the most recent graphic that the GAM with 4 knots is unable to capture a necessary amonunt of the variation present in our data.