---
title: "Assignment 11; STAT 689"
author: "Philip Anderson; panders2@tamu.edu"
date: "4/02/2018"
output: pdf_document
---

```{r, message=FALSE}
library("HRW")
library("mgcv")
library("cosso")
library("tidyverse")
```

# Question 1

```{r}
data("BostonMortgages")
```

```{r}
str(BostonMortgages)
```

## 1A
Which model does Cosso select?

```{r}
# create a 1/0 indicator for deny so we know what the procedures are predicting
BostonMortgages$deny_bin <- ifelse(BostonMortgages$deny=="yes", 1, 0)
# evaluate results with contingency table
table(BostonMortgages$deny_bin, BostonMortgages$deny)
```

```{r}
# generate design matrix
# column 5 has a huge outlier - drop it
BostonMortgages_red <- BostonMortgages %>%
  dplyr::filter(dir < 1)

# one-hot-encode the factor variables and drop the intercept this function makes
fac_to_int <- stats::model.matrix(~ black + pbcr + self + single, BostonMortgages_red)[, -1] %>%
  data.frame()
int_vars <- BostonMortgages_red[, c("dir", "lvr")]
X <- as.matrix(cbind(fac_to_int, int_vars))

# generate response array
y <- BostonMortgages_red[, c("deny_bin")]
```


```{r}
start <- Sys.time()
cosso_mod <- cosso::cosso(x=X, y=y, family=c("Binomial"))
end <- Sys.time()
# runtime
print(end - start)
```
```{r}
head(X)
```

```{r}
tune_matrix <- data.frame(cosso_mod$tune$Mgrid
                  , cosso_mod$tune$L2norm
                )
names(tune_matrix) <- c("M","blackyes", "pbcryes", "selfyes", "singleyes", "dir", "lvr")
print(round(tune_matrix, 2))
```

```{r}
start <- Sys.time()
tune_mod <- cosso::tune.cosso(cosso_mod, 4, FALSE)
fin <- Sys.time()
print(fin - start)
```

```{r}
cosso::plot.cosso(cosso_mod, M=tune_mod$OptM)
```

```{r}
tune_matrix %>% 
  dplyr::filter(M==tune_mod$OptM) %>%
  print()
```

The only variable selected for our model is _dir_. 

## 1B
Which model does mgcv select?

I am going to begin by throwing a great deal of information into the model, using _deny_ as the outcome of interest.


```{r}
# use an equivalent model to the one above - continue using the Boston Mortgages dataset
mgcv_mod <- mgcv::gam(deny_bin ~ black + pbcr + self + single + dir + lvr
                      , data=BostonMortgages_red
                      , family=binomial(link="logit")
                      , select=T
                      )
mgcv::summary.gam(mgcv_mod)
```

The mgcv procedure selected all of the variables that we put into the model.

## 1C
The variables selected are not close.  Reconciliation between the opposed results would require fairly extensive troubleshooting that will not be performed here.

# Question 2

```{r}
data("femSBMD")
names(femSBMD) <- tolower(names(femSBMD))
str(femSBMD)
```
## 2A
Define a new identification variable, given as 2 times the existing identification variable.

```{r}
femSBMD$idnum2 <- 2*femSBMD$idnum
str(femSBMD)
```

## 2B
Rerun the gamm model given in class, using the original idnum variable and the new one.  Ensure that we are getting the same results with each.

```{r}
# original fit
class_fit <- mgcv::gamm(spnbmd ~ s(age) + black + hispanic + white
                        , random=list(idnum = ~1)
                        , data=femSBMD
                         )
# new fit 
hw_fit <- mgcv::gamm(spnbmd ~ s(age) + black + hispanic + white
                        , random=list(idnum2 = ~1)
                        , data=femSBMD
                         )
```

```{r}
# print the original results
summary(class_fit$gam)
```

```{r}
# print out the updated results
summary(hw_fit$gam)
```

The specific number of the id number variable does not appear to matter in the gamm function's grouping of observations.



# Question 3
```{r}
(version_info <- version)

version_num <- paste0(version_info$major, ".", version_info$minor)
cat("\n")
print(paste0("version num: ", version_num))
```

```{r}
library("rstan")
```

RStan has loaded successfully.


# Question 4

```{r}
pig.weights <- read.csv("/Users/panders2/Documents/schools/tamu/stat_689/homework/semiparametric-regression/hw11/pigWeights.csv")
names(pig.weights) <- tolower(names(pig.weights))
str(pig.weights)
```

```{r, message=F}
library("lattice")
```

```{r}
lattice::xyplot(weight~num.weeks
                , data=pig.weights
                , groups=id.num
                , type="b"
                 )
```

## 4B
Looking at the data, it appears that a random-intercept model holds for these data.  Note that the slope of each pig's data appears to be the same, and they tend to end up where they started.

## 4C
Fit the random-intercept model and give your code.  Also, do a summary and show your results.

```{r}
random_int <- mgcv::gamm(weight ~ num.weeks
                         , random=list(id.num = ~1)
                         , data=pig.weights
                         )
summary(random_int$gam)
```


