---
title: "Assignment 16; STAT 689"
author: "Philip Anderson; panders2@tamu.edu"
date: "4/30/2018"
output: pdf_document
---

```{r setup, include=FALSE}
library("tidyverse")
library("mgcv")
library("quantreg")
```

```{r}
# bring in data
sim <- read.csv("/Users/panders2/Documents/schools/tamu/stat_689/homework/semiparametric-regression/misc/OPEN_sim.csv")
names(sim) <- tolower(names(sim))
str(sim)
```

```{r}
# response
y <- sim$bio
# smoothed
x1 <- sim$avgffq
x2 <- sim$avgrecall
# linear
z1 <- sim$age
z2 <- sim$bmi
```


# Question 1
Fit a quantile regression with tau=0.5 for the regression of Y on X1 and Z.

```{r}
mod_one <- quantreg::rqss(y ~ qss(x1, lambda=3.5) + z1 + z2
                          , tau=0.5 # modeling the median
                          )
```

## Question 1A
What is statistically significant?

```{r}
summary(mod_one)
```

The smoothed x1 term, or avgFFQ is the only significant term at the 5% level.

## Question 1B
Plot the fitted function when age=55 and BMI=25.

```{r}
ng <- 101
ffqg <- seq(from=min(x1)
            , to=max(x1)
            , length=ng
            )
newData <- as.data.frame(cbind(ffqg
              , rep(55, ng)
              , rep(25, ng) 
                    )
              )
names(newData) <- c("x1", "z1", "z2")

newDataList <- as.list(newData)

predObj <- quantreg::predict.rqss(object=mod_one
                   , newdata=newDataList
                   , interval="confidence"
                   , level=0.95
                   )
```

```{r}
plot(ffqg, predObj[,1], type="n")
polygon(c(ffqg, rev(ffqg))
        , c(predObj[,2], rev(predObj[,3]))
        , col="cadetblue1"
        , border=F
        )
lines(ffqg, predObj[,1], lwd=2)
rug(x1)
```



# Question 2
Fit an ordinary regression, but apply a bivariate spline, te(x1, x2).

```{r}
mod_two <- mgcv::gam(y ~ te(x1, x2) + z1 + z2
                     , method="REML"
                     )
```

## Question 2A
What is statistically significant?

```{r}
summary(mod_two)
```

Of the linear terms, only z2 (BMI) is significant; the bivariate smoothed term of x1 and x2 is also significant.

## Question 2B
Plot the functions when age=55 and BMI=25.  

X1 materials first.
```{r}
ng <- 101
x1g <- seq(from=min(x1), to=max(x1), length=ng)

newDataDF <- as.data.frame(
          cbind(
          x1g
          , rep(mean(x2), ng)
          , rep(55, ng)
          , rep(25, ng)
          )
  )
names(newDataDF) <- c("x1", "x2", "z1", "z2")
newDataList <- as.list(newDataDF)

predObj_one <- predict(mod_two, newDataList, se=T)

muHatg_one <- predObj_one$fit
aa_one     <- predObj_one$fit + 2*predObj_one$se
bb_one     <- predObj_one$fit - 2*predObj_one$se
lowergg_one <- 1 / (1 + exp(-bb_one)) 
uppergg_one <- 1 / (1 + exp(-aa_one)) 

```


X2 materials second.


```{r}
ng <- 101
x2g <- seq(from=min(x2), to=max(x2), length=ng)

newDataDF <- as.data.frame(
          cbind(
          rep(mean(x1), ng)
          , x2g
          , rep(55, ng)
          , rep(25, ng)
          )
  )
names(newDataDF) <- c("x1", "x2", "z1", "z2")
newDataList <- as.list(newDataDF)

predObj_two <- predict(mod_two, newDataList, se=T)

muHatg_two <- predObj_two$fit
aa_two     <- predObj_two$fit + 2*predObj_two$se
bb_two     <- predObj_two$fit - 2*predObj_two$se
lowergg_two <- 1 / (1 + exp(-bb_two)) 
uppergg_two <- 1 / (1 + exp(-aa_two)) 

```

```{r}
plot(x1g, muHatg_one, type="n")
polygon(c(x1g, rev(x1g))
       , c(lowergg_one, rev(uppergg_one))
       , col="cadetblue1"
        )
```
