---
title: "Assignment 12; STAT 689"
author: "Philip Anderson; panders2@tamu.edu"
date: "4/5/2018"
output: pdf_document
---

```{r, message=F}
library("HRW")
library("mgcv")
library("nlme")
library("lattice")
library("tidyverse")
```

```{r}
# bring in the data
pigs <- read.csv('/Users/panders2/Documents/schools/tamu/stat_689/homework/semiparametric-regression/hw11/pigWeights.csv')
str(pigs)
```

# Question 1
## 1A 
How many pigs are in the model?

```{r}
cat('Distinct pigs in the data given by: ', length(unique(pigs$id.num)))
```

## 1B
Fit the random function model and display your code.  I am going to model pig weight as a function of the number of weeks since measurement on the pigs began.

```{r}
# extract the important variables into individual objects
id_num <- pigs$id.num
num_weeks <- pigs$num.weeks
pig_weight <- pigs$weight
```

Now, we need to set up the design matrices for the splines at the population (global) level and individual (group) level.  Note that the individual level will have fewer knots than the population level.  We will need to follow this with a random effects structure.  There is a lot of code here that will be reused later, so I am going to package this up into a function - note that it won't be generalizable beyond this specific situation.


```{r}
random_func_mod <- function(id_num, num_weeks, pig_weight) {
  
      # number records
      numObs <- length(id_num)
      # number of subjects
      numGrp <- length(unique(id_num))
      
      ## population (Gbl) work
      # knots
      numIntKnotsGbl <- 20
      # population O-Sull Basis Functions 
      intKnotsGbl <- quantile(unique(num_weeks)
                              , seq(0, 1, length=numIntKnotsGbl+2)
                              )[-c(1, numIntKnotsGbl+2)]
      range.num_weeks <- c(min(num_weeks)-0.01, max(num_weeks)+0.01)
      
      Zgbl <- HRW::ZOSull(num_weeks, range.x=range.num_weeks, intKnots=intKnotsGbl)
      
      ## subject-level (Grp) work
      numIntKnotsGrp <- 10
      intKnotsGrp <- quantile(unique(num_weeks)
                              , seq(0, 1, length=numIntKnotsGrp+2)
                              )[-c(1, numIntKnotsGrp+2)]
      Zgrp <- HRW::ZOSull(num_weeks, range.x=range.num_weeks, intKnots=intKnotsGrp)
      
      
      #Now, set up the random effects structure.
      dummyId <- factor(rep(1, numObs))
      Zblock <- list(
                    dummyId = pdIdent( ~ -1 + Zgbl) 
                    , id_num = pdSymm(~ num_weeks)
                    , id_num = pdIdent(~ -1 + Zgrp)
                    )
      gd <- groupedData(pig_weight ~ num_weeks|rep(1, length=numObs)
                        , data=data.frame(pig_weight, num_weeks, Zgbl, Zgrp, id_num)
                        )
      fit <- lme(pig_weight ~ num_weeks, data=gd, random=Zblock)
      
      return(fit)
}


```

```{r}
fit <- random_func_mod(id_num=id_num, num_weeks=num_weeks, pig_weight=pig_weight)

```

Display the summary statistics that are mentioned in Lecture 18.

```{r}
summary(fit)
```

# Question 2
Plot the population-level BLUP estimates.

```{r}
# number of grid points
ng <- 101
# grid for num_weeks
num_weeks_g <- seq(range.num_weeks[1], range.num_weeks[2], length=ng)
# design matrix for linear component; col of 1's plus num_weeks grid
Xg <- cbind(rep(1, ng), num_weeks_g)

# spline terms - overall fit
Zgblg <- HRW::ZOSull(num_weeks_g, range.x=range.num_weeks, intKnots=intKnotsGbl)
# spline terms for individual fits
Zgrpg <- HRW::ZOSull(num_weeks_g, range.x=range.num_weeks, intKnots=intKnotsGrp)

# grab betaHat, the model intercept, and the slope from our model objet
betaHat <- as.vector(fit$coefficients$fixed)
# grab uHat, along with the estimated spline coef for overall fit
uHat <- as.vector(fit$coefficients$random[[1]])
# form the overall fit
fHatg <- as.vector(Xg %*% betaHat + Zgblg %*% uHat)
# subject-specific estimated curves
curvEsts <- vector("list", numGrp)

for (i in 1:numGrp)
  {
  # subject-specific slope + intercept
  uLinHati <- as.vector(fit$coefficients$random[[2]][i, ])
  # subject-specific terms for splines
  uSplHati <- as.vector(fit$coefficients$random[[3]][i, ])
  # individual function estimates
  ghati <- Xg %*% uLinHati + Zgrpg %*% uSplHati
  curvEsts[[i]] <- fHatg + ghati
  }

```


```{r}
plot(num_weeks_g, fHatg, type="l", col="dodgerblue4", lwd=3
     , xlab="Number of Weeks" 
     , ylab="Weight"
     , main="Pig Weight by Number of Weeks - Population Curve"
     )
```

# Question 3
Plot the individual BLUP estimates for the first 28 pigs.

```{r}
# reduce the data to just the first 28 pigs
pigs_28 <- pigs %>%
  dplyr::filter(id.num <= 28)

id_num_28 <- pigs_28$id.num
num_weeks_28 <- pigs_28$num.weeks
pig_weight_28 <- pigs_28$weight
```

```{r}
figure <- xyplot(pig_weight_28 ~ num_weeks_28 | id_num_28
                 , groups=id_num_28
                 , strip=F
                 , scales=list(cex=1.25)
                 , xlab=list("age (years)", cex=1.5)
                 , ylab=list("height (centimeters)", cex=1.5)
                 , as.table=T
                 , main="Pig Weight for first 28 Pigs"
#                 , layout=c(4,7)
                 , panel=function(x, y, subscripts, groups)
                     {
                        panel.grid()
                        adolNum  <- id_num_28[subscripts][1]
                        panel.superpose(x, y, subscripts, groups
                                        , col="dodgerblue", type="b"
                                        )
                        panel.xyplot(num_weeks_g, curvEsts[[adolNum]]
                                     , col="blue", type="l"
                                     )
                     }
                 )
figure

```

# Question 4
Rerun problem 1 for separate time periods, first weeks 1-5, and then for weeks 6-9.

```{r}
pigs <- read.csv('/Users/panders2/Documents/schools/tamu/stat_689/homework/semiparametric-regression/hw11/pigWeights.csv')
# first split out the data
pigs_epoch_1 <- pigs %>%
  dplyr::filter(num.weeks < 6)
pigs_epoch_2 <- pigs %>%
  dplyr::filter(num.weeks >= 6)

```

First, weeks 1-5.
```{r}
id_num <- pigs_epoch_1$id.num
num_weeks <- pigs_epoch_1$num.weeks
pig_weight <- pigs_epoch_1$weight

fit_epoch_1 <- random_func_mod(id_num=id_num, num_weeks=num_weeks, pig_weight=pig_weight)
```


Now, weeks 6-9.
```{r}
id_num <- pigs_epoch_2$id.num
num_weeks <- pigs_epoch_2$num.weeks
pig_weight <- pigs_epoch_2$weight

fit_epoch_2 <- random_func_mod(id_num=id_num, num_weeks=num_weeks, pig_weight=pig_weight)
```

```{r}
stop('stop')
```

I am going to work with the first time period first.  Because a lot of this code is recycled from above, I will have reduced comments.

```{r}
# extract meaningful variables into objects
id_num <- pigs_epoch_1$id.num
num_weeks <- pigs_epoch_1$num.weeks
pig_weight <- pigs_epoch_1$weight
```

```{r}
# set up our design matrices.
# number records
numObs <- length(id_num)
# number of subjects
numGrp <- length(unique(id_num))

## population (Gbl) work
# knots
numIntKnotsGbl <- 20
# population O-Sull Basis Functions 
intKnotsGbl <- quantile(unique(num_weeks)
                        , seq(0, 1, length=numIntKnotsGbl+2)
                        )[-c(1, numIntKnotsGbl+2)]
range.num_weeks <- c(min(num_weeks)-0.01, max(num_weeks)+0.01)

Zgbl <- HRW::ZOSull(num_weeks, range.x=range.num_weeks, intKnots=intKnotsGbl)

## subject-level (Grp) work
numIntKnotsGrp <- 10
intKnotsGrp <- quantile(unique(num_weeks)
                        , seq(0, 1, length=numIntKnotsGrp+2)
                        )[-c(1, numIntKnotsGrp+2)]
Zgrp <- HRW::ZOSull(num_weeks, range.x=range.num_weeks, intKnots=intKnotsGrp)
```

```{r}
# random effects structure
dummyId <- factor(rep(1, numObs))
Zblock <- list(
              dummyId = pdIdent( ~ -1 + Zgbl) 
              , id_num = pdSymm(~ num_weeks)
              , id_num = pdIdent(~ -1 + Zgrp)
              )
gd <- groupedData(pig_weight ~ num_weeks|rep(1, length=numObs)
                  , data=data.frame(pig_weight, num_weeks, Zgbl, Zgrp, id_num)
                  )
fit_epoch_1 <- lme(pig_weight ~ num_weeks, data=gd, random=Zblock)
```

Now going to work with the second period.


```{r}
# extract meaningful variables into objects
id_num <- pigs_epoch_2$id.num
num_weeks <- pigs_epoch_2$num.weeks
pig_weight <- pigs_epoch_2$weight
```

```{r}
# set up our design matrices.
# number records
numObs <- length(id_num)
# number of subjects
numGrp <- length(unique(id_num))

## population (Gbl) work
# knots
numIntKnotsGbl <- 20
# population O-Sull Basis Functions 
intKnotsGbl <- quantile(unique(num_weeks)
                        , seq(0, 1, length=numIntKnotsGbl+2)
                        )[-c(1, numIntKnotsGbl+2)]
range.num_weeks <- c(min(num_weeks)-0.01, max(num_weeks)+0.01)

Zgbl <- HRW::ZOSull(num_weeks, range.x=range.num_weeks, intKnots=intKnotsGbl)

## subject-level (Grp) work
numIntKnotsGrp <- 10
intKnotsGrp <- quantile(unique(num_weeks)
                        , seq(0, 1, length=numIntKnotsGrp+2)
                        )[-c(1, numIntKnotsGrp+2)]
Zgrp <- HRW::ZOSull(num_weeks, range.x=range.num_weeks, intKnots=intKnotsGrp)
```

```{r}
# random effects structure
dummyId <- factor(rep(1, numObs))
Zblock <- list(
              dummyId = pdIdent( ~ -1 + Zgbl) 
              , id_num = pdSymm(~ num_weeks)
              , id_num = pdIdent(~ -1 + Zgrp)
              )
gd <- groupedData(pig_weight ~ num_weeks|rep(1, length=numObs)
                  , data=data.frame(pig_weight, num_weeks, Zgbl, Zgrp, id_num)
                  )
fit_epoch_2 <- lme(pig_weight ~ num_weeks, data=gd, random=Zblock)
```

Now, let's compare the output of these models.

```{r}
fit_epoch_1$sigma
fit_epoch_2$sigma
```