---
title: "Assignment 12; STAT 689"
author: "Philip Anderson; panders2@tamu.edu"
date: "4/10/2018"
output: pdf_document
---

```{r, message=F}
# preliminaries
rm(list=ls())
library("HRW")
library("mgcv")
library("nlme")
library("lattice")
library("tidyverse")
```

```{r}
# bring in the data
pigs <- read.csv('/Users/panders2/Documents/schools/tamu/stat_689/homework/semiparametric-regression/hw11/pigWeights.csv')
str(pigs)
```

# Question 1
## 1A 
How many pigs are in the model?

```{r}
cat('Distinct pigs in the data given by: ', length(unique(pigs$id.num)))
```

## 1B
Fit the random function model and display your code.  
I am going to model pig weight as a function of the number of weeks since measurement on the pigs began.

```{r}
# extract the important variables into individual objects
id_num_a <- pigs$id.num
num_weeks_a <- pigs$num.weeks
pig_weight_a <- pigs$weight
```

Now, we need to set up the design matrices for the splines at the population (global) level and individual (group) level.  Note that the individual level will have fewer knots than the population level.  We will follow this with a random effects structure, generated below.  There is a lot of code here that will be reused later, so I am going to package this up into a function - note that it won't be generalizable beyond this specific situation.


```{r}
random_func_mod <- function(id_num, num_weeks, pig_weight) {
  
      # a lot of these objects get used down the road, so I am going to use global assignment
      
      # number of records
      numObs <<- length(id_num)
      
      # number of subjects
      numGrp <<- length(unique(id_num))
      
      # population (Gbl) work
      # knots
      numIntKnotsGbl <<- 20
      # population O-Sull Basis Functions 
      intKnotsGbl <<- quantile(unique(num_weeks)
                              , seq(0, 1, length=numIntKnotsGbl+2)
                              )[-c(1, numIntKnotsGbl+2)]
      range.num_weeks <<- c(min(num_weeks)-0.01, max(num_weeks)+0.01)
      
      Zgbl <<- HRW::ZOSull(num_weeks, range.x=range.num_weeks, intKnots=intKnotsGbl)
      
      # subject-level (Grp) work
      numIntKnotsGrp <<- 10
      intKnotsGrp <<- quantile(unique(num_weeks)
                              , seq(0, 1, length=numIntKnotsGrp+2)
                              )[-c(1, numIntKnotsGrp+2)]
      Zgrp <<- HRW::ZOSull(num_weeks, range.x=range.num_weeks, intKnots=intKnotsGrp)
      
      
      #Now, set up the random effects structure.
      dummyId <<- factor(rep(1, numObs))
      Zblock <<- list(
                    dummyId = pdIdent( ~ -1 + Zgbl) 
                    , id_num = pdSymm(~ num_weeks)
                    , id_num = pdIdent(~ -1 + Zgrp)
                    )
      gd <<- groupedData(pig_weight ~ num_weeks|rep(1, length=numObs)
                        , data=data.frame(pig_weight, num_weeks, Zgbl, Zgrp, id_num)
                        )
      fit <<- lme(pig_weight ~ num_weeks, data=gd, random=Zblock)
      
      return(fit)
      
}


```

```{r}
fit <- random_func_mod(id_num=id_num_a, num_weeks=num_weeks_a, pig_weight=pig_weight_a)
```

Display the summary.

```{r}
summary(fit)
```

# Question 2
Plot the population-level BLUP estimates.  First, we need to work through a number of preliminaries.

```{r}
# number of grid points
ng <- 101
# grid for num_weeks
num_weeks_g <- seq(range.num_weeks[1], range.num_weeks[2], length=ng)
# design matrix for linear component; col of 1's plus num_weeks grid
Xg <- cbind(rep(1, ng), num_weeks_g)

# spline terms - overall fit
Zgblg <- HRW::ZOSull(num_weeks_g, range.x=range.num_weeks, intKnots=intKnotsGbl)
# spline terms for individual fits
Zgrpg <- HRW::ZOSull(num_weeks_g, range.x=range.num_weeks, intKnots=intKnotsGrp)

# grab betaHat, the model intercept, and the slope from our model objet
betaHat <- as.vector(fit$coefficients$fixed)
# grab uHat, along with the estimated spline coef for overall fit
uHat <- as.vector(fit$coefficients$random[[1]])
# form the overall fit
fHatg <- as.vector(Xg %*% betaHat + Zgblg %*% uHat)
# subject-specific estimated curves
curvEsts <- vector("list", numGrp)

for (i in 1:numGrp)
  {
  # subject-specific slope + intercept
  uLinHati <- as.vector(fit$coefficients$random[[2]][i, ])
  # subject-specific terms for splines
  uSplHati <- as.vector(fit$coefficients$random[[3]][i, ])
  # individual function estimates
  ghati <- Xg %*% uLinHati + Zgrpg %*% uSplHati
  curvEsts[[i]] <- fHatg + ghati
  }

```

Now do the population-level plot.
```{r}
plot(num_weeks_g, fHatg, type="l", col="dodgerblue4", lwd=3
     , xlab="Number of Weeks" 
     , ylab="Weight"
     , main="Pig Weight by Number of Weeks - Population Curve"
     )
```
Overall, this plot looks very linear.

# Question 3
Plot the individual BLUP estimates for the first 28 pigs.

```{r}
# reduce the data to just the first 28 pigs
pigs_28 <- pigs %>%
  dplyr::filter(id.num <= 28)

id_num_28 <- pigs_28$id.num
num_weeks_28 <- pigs_28$num.weeks
pig_weight_28 <- pigs_28$weight
```

```{r}
figure <- xyplot(pig_weight_28 ~ num_weeks_28 | id_num_28
                 , groups=id_num_28
                 , strip=F
                 , scales=list(cex=1.25)
                 , xlab=list("age (years)", cex=1.5)
                 , ylab=list("height (centimeters)", cex=1.5)
                 , as.table=T
                 , main="Pig Weight for n=1:28"
#                 , layout=c(4,7)
                 , panel=function(x, y, subscripts, groups)
                     {
                        panel.grid()
                        adolNum  <- id_num_28[subscripts][1]
                        panel.superpose(x, y, subscripts, groups
                                        , col="dodgerblue", type="b"
                                        )
                        panel.xyplot(num_weeks_g, curvEsts[[adolNum]]
                                     , col="blue", type="l"
                                     )
                     }
                 )
figure

```


# Question 4
Rerun problem 1 for separate time periods, first weeks 1-5, and then for weeks 6-9.

```{r}
# first split out the data
pigs_epoch_1 <- pigs %>%
  dplyr::filter(num.weeks < 6)
pigs_epoch_2 <- pigs %>%
  dplyr::filter(num.weeks >= 6)

```

First, weeks 1-5.
```{r}
id_num_1 <- pigs_epoch_1$id.num
num_weeks_1 <- pigs_epoch_1$num.weeks
pig_weight_1 <- pigs_epoch_1$weight

fit_epoch_1 <- random_func_mod(id_num=id_num_1, num_weeks=num_weeks_1, pig_weight=pig_weight_1)
```


Now, weeks 6-9.
```{r}
id_num_2 <- pigs_epoch_2$id.num
num_weeks_2 <- pigs_epoch_2$num.weeks
pig_weight_2 <- pigs_epoch_2$weight

fit_epoch_2 <- random_func_mod(id_num=id_num_2, num_weeks=num_weeks_2, pig_weight=pig_weight_2)
```

Print General Summaries of each fit.
```{r}
summary(fit_epoch_1)
```

```{r}
summary(fit_epoch_2)
```

Looking specifically at the between-subject variability for these time periods:

```{r}
print("Weeks 1-5 between-subject variability")
print(fit_epoch_1$modelStruct$reStruct[3])
print("Weeks 6-9 between-subject variability")
print(fit_epoch_2$modelStruct$reStruct[3])
```

I find the between-subject variability to be greater for model 1, but not dramatically.
