---
title: "Assignment-07"
author: "Philip Anderson"
date: "2/22/2018"
output: pdf_document
---

```{r setup, message=FALSE}
library("mgcv")
```

```{r}
# read it in
fram <- read.csv("/Users/panders2/Documents/schools/tamu/stat_689/homework/semiparametric-regression/misc/framingham_with_lsbp_and_lcholest.csv")
names(fram) <- tolower(names(fram))
str(fram)
```

Our response variable Y, equals chd; our predictors are lsbp, lcholest, age, and smoking status.  

# Question 1
Fit a logistic gam with only LSBP modeled within a spline function.

```{r}
logit1 <- mgcv::gam(chd ~ s(lsbp, k=23, bs="cr") + lcholest + age + smoker
                    , family=binomial(link="logit")
                    , data=fram
                    )
summary(logit1)
```

P-values for 4 predictors are as follows:

```{r}
pval_smry <- function(gam_mod) {
  gam_smry <- summary(gam_mod)
  print("Linear Term p-values:")
  print(gam_smry$pTerms.pv)
  cat("\nSmoothed Term p-values:\n")
  cat(gam_smry$s.pv)
  
}

pval_smry(logit1)
```

From the p-value on the smoothed term, it appears that lsbp should be modeled as a spline.

# Question 2
Fit a logisitc gam with only lcholest modeled as a spline.

```{r}
logit2 <- mgcv::gam(chd ~ lsbp + s(lcholest, k=23, bs="cr") + age + smoker
                    , family=binomial(link="logit")
                    , data=fram
                    )
summary(logit2)

```

```{r}
pval_smry(logit2)
```

It looks as though the _lcholest_ variable should be modeled within a spline function.

# Question 3
Fit a logistic gam with only age modeled as a spline.

```{r}
logit3 <- mgcv::gam(chd ~ lsbp + lcholest + s(age, k=23, bs="cr") + smoker
                    , family=binomial(link="logit")
                    , data=fram
                    )
summary(logit3)
``` 

```{r}
pval_smry(logit3)
```

It appears as though the age term should be smoothed as well.

# Question 4
Fit the model with all continuous terms modeled as splines.

```{r}
logit4 <- mgcv::gam(chd ~ s(lsbp, k=23, bs="cr") + 
                      s(lcholest, k=23, bs="cr") + 
                      s(age, k=23, bs="cr") + smoker
                    , family=binomial(link="logit")
                    , data=fram
                    )
summary(logit4)

```

```{r}
pval_smry(logit4)
```

# Question 5
Refit the model from Question 4, with only the terms that you picked as being significant under spline transformation.

```{r}
logit_alt <- mgcv::gam(chd ~ lsbp + lcholest + s(age) + smoker
                       , family=binomial(link="logit")
                       , data=fram
                      )
```



```{r}
# specify the null model
logit_null <- mgcv::gam(chd ~ lsbp + lcholest + age + smoker
                        , family=binomial(link="logit")
                        , data=fram
                      )
```


```{r}
anova(logit_alt, logit_null, test="Chisq")
```

The moel that contains the smoothed terms is not significantly different from the model that doesn't.

```{r}

print_em <- function(mod_title="Model 1", mod_obj) {

  print(mod_title)
  mod_smry <- summary(mod_obj)
  print("Smoke impact odds ratio:")
  print(mod_smry$p.coeff[3 + 1])
  print("Smoke p-value:")
  print(mod_smry$pTerms.pv[3])
  cat("\n--- --- --- --- --- \n\n")
}

print_em(mod_title="Model 1", mod_obj=logit1)
print_em(mod_title="Model 2", mod_obj=logit2)
print_em(mod_title="Model 3", mod_obj=logit3)
print_em(mod_title="Model 4", mod_obj=logit4)


```








