---
title: "Assignment10; STAT 689"
author: "Philip Anderson; panders2@tamu.edu"
date: "3/21/2018"
output: pdf_document
---

# Question 10
## 10A
```{r, message=FALSE}
library("mgcv")
library("HRW")
data(ragweed)
str(ragweed)
```

## 10B

```{r}
fit1 <- mgcv::gam(pollenCount ~ factor(year) + s(dayInSeason, k=27, by=factor(year))
                  + temperatureResidual + rain + s(windSpeed, k=27) 
                 , data=ragweed
                 , family=poisson
                   )

```

Obtain residual plots for the models.
```{r}
mgcv::gam.check(fit1)
qqnorm(residuals(fit1, type="scaled.pearson"), main="Second QQ-Plot")
abline(0, 1, col="red")
```

## 10C
Obtain alternative fits.

```{r}
# quasipoisson
fit2 <- mgcv::gam(pollenCount ~ factor(year) + s(dayInSeason, k=27, by=factor(year))
                  + temperatureResidual + rain + s(windSpeed, k=27) 
                 , data=ragweed
                 , family=quasipoisson
                   )

```

```{r}
# square root variance transformation
fit3 <- mgcv::gam(sqrt(pollenCount) ~ factor(year) + s(dayInSeason, k=27, by=factor(year))
                  + temperatureResidual + rain + s(windSpeed, k=27) 
                 , data=ragweed
                 , family=gaussian(link="identity")
                   )

```

## 10D
Obtain residual plots for the new fits, analogous to those obtained for fit1 in Part B

```{r}
# quasipoisson
mgcv::gam.check(fit2)
qqnorm(residuals(fit2, type="scaled.pearson")
       , main="Quasipoisson Fit: Second QQ-Plot")
abline(0, 1, col="red")
```

```{r}
mgcv::gam.check(fit3)
qqnorm(residuals(fit3, type="scaled.pearson")
       , main="Gaussian Fit: Second QQ-Plot"
       )
abline(0, 1, col="red")
```

## 10E

Based on the above summaries, I believe that the Gaussian fit for the square root transformation (fit3) is superior to the Poisson and QuasiPoisson fits.  The scaled Pearson residual QQ-Norm plot shows that the Gaussian response distribution seems to result in a superior fit to the data.

```{r}
# graphical representations are available in the previous question
# numerical summary immediately below.
mgcv::summary.gam(fit3)
```

## 10F
Show the four penalized spline fits for the effect of dayInSeason on the mean response.

```{r}
par(mfrow=c(2,2))
plot(fit3, shade=TRUE, col="blue", rug=TRUE, select=c(1)
     , main="dayInSeason for 1991"
     , ylab="pollen count"
     )
plot(fit3, shade=TRUE, col="blue", rug=TRUE, select=c(2)
     , main="dayInSeason for 1992"
     , ylab="pollen count"
     )
plot(fit3, shade=TRUE, col="blue", rug=TRUE, select=c(3)
     , main="dayInSeason for 1993"
     , ylab="pollen count"
     )
plot(fit3, shade=TRUE, col="blue", rug=TRUE, select=c(4)
     , main="dayInSeason for 1994"
     , ylab="pollen count"
     )
par(mfrow=c(1,1))
```

# Question 2

For the square root transformation (fit3), see whether the idea of constant variance is reasonable.
```{r}
resid <- abs(residuals(fit3))
fitted <- fitted(fit3)
plot(fitted, resid
     , main="Fitted Values vs. Absolute Residuals - Fit 3"
     , xlab="Fitted Values"
     , ylab="Abs. Residuals"
     )

```

From the above plot, it does not appear that the assumption of constant variance is reasonable.  As our fitted values increase, the variance of our absolute residuals increases.  

# Question 3
From the estimated scale in the QuasiPoisson model, does the scale parameter imply that we may have concern about overdispersion?

```{r}
print("Scale Parameter given by:")
print(fit2$scale)
```

Considering that we are looking for a scale value of 1, I believe we do have legitimate concerns about overdispersion.

## Question 4
Test whether the QuasiPoisson fit is statistically significantly different from the Poisson fit.

```{r}
anova(fit1, fit2, test="Chisq")
```

It appears that we have statistically significant differences between the Poisson and QuasiPoisson fits.  The QuasiPoisson is preferable in this case.

# Question 5
In the Poisson fit, is there evidence that there is a year by dayInSeason interaction?

```{r}
mgcv::summary.gam(fit1)
```

From the above output, it appears that we have a significant interaction between the year and dayInSeason variables. 

We can also create similar model that does not have the interaction, and test them against each other.
```{r}
smaller_fit1 <- mgcv::gam(pollenCount ~ factor(year) + s(dayInSeason, k=27)
                  + temperatureResidual + rain + s(windSpeed, k=27) 
                 , data=ragweed
                 , family=poisson
                   )
anova(fit1, smaller_fit1, test="Chisq")

```

Both evaluations suggest significant evidence of a year by dayInSeason interaction.

# Question 6
Remove the interaction from the model, still doing the quasipoisson; do a stepwise regression - which model is seleted?

```{r, message=FALSE, error=T}
detach("package:mgcv", unload=TRUE)
library("gam")
```

For the following, the quasipoisson response distribution appears to be incompatible with the step.gam function.  I use poisson and gaussian response distributions for model selection.
```{r}
pois_init <- gam::gam(pollenCount ~ factor(year) + dayInSeason
                  + temperatureResidual + rain + windSpeed
                 , data=ragweed
                 , family=poisson(link="log")
                   )


pois_step <- gam::step.gam(pois_init, scope=list(
                        "year" = ~1 + year
                        , "dayInSeason" = ~1 + dayInSeason + s(dayInSeason, 2)
                        , "temperatureResidual" = ~1 + temperatureResidual 
                        , "rain" = ~1 + rain
                        , "windSpeed" = ~1 + windSpeed + s(windSpeed, 2)
                        )
                          )
```

```{r}
gauss_init <- gam::gam(sqrt(pollenCount) ~ factor(year) + dayInSeason
                  + temperatureResidual + rain + windSpeed
                 , data=ragweed
                 , family=gaussian(link="identity")
                   )


gauss_step <- gam::step.gam(gauss_init, scope=list(
                        "year" = ~1 + year
                        , "dayInSeason" = ~1 + dayInSeason + s(dayInSeason, 2)
                        , "temperatureResidual" = ~1 + temperatureResidual 
                        , "rain" = ~1 + rain
                        , "windSpeed" = ~1 + windSpeed + s(windSpeed, 2)
                        )
                          )
```

Both GLMs result in the same model:

```{r}
names(gauss_step$model)[-1]
```


# Question 7
Fit a model using cosso.  COSSO does not have the Poisson or QuasiPoisson response distributions available, so I will use Guassian and transform the response in some of the models.
```{r}
Y <- ragweed$pollenCount
Y_alt <- log(ragweed$pollenCount)
Y_alt2 <- sqrt(ragweed$pollenCount)
X <- ragweed[, c(2, 3, 5, 6, 7)]
# make sure everything is in 'integer format'
X$temperatureResidual <- as.integer(X$temperatureResidual)
X$windSpeed <- as.integer(X$windSpeed)
str(X)
```

```{r, message=FALSE, error=T}
cosso_mod1 <- cosso::cosso(x=X, y=Y, family="Gaussian")
```
```{r, message=FALSE, error=T}
cosso_mod2 <- cosso::cosso(x=X, y=Y_alt, family="Gaussian")
```

```{r, message=FALSE, error=T}
cosso_mod3 <- cosso::cosso(x=X, y=Y_alt2, family="Gaussian")
```

None of the cosso models were able to run successfully.
