---
title: "Assignment10; STAT 689"
author: "Philip Anderson; panders2@tamu.edu"
date: "3/21/2018"
output: pdf_document
---

# Question 10
## 10A
```{r}
library("HRW")
data(ragweed)

```

```{r}
str(ragweed)
```

## 10B

```{r}
fit1 <- mgcv::gam(pollenCount ~ factor(year) + s(dayInSeason, k=27, by=factor(year))
                  + temperatureResidual + rain + s(windSpeed, k=27) 
                 , data=ragweed
                 , family=poisson
                   )

```


```{r}
mgcv::gam.check(fit1)
qqnorm(residuals(fit1, type="scaled.pearson"))
abline(0, 1, col="red")
```

## 10C
Obtain alternative fits

```{r}
fit2 <- mgcv::gam(pollenCount ~ factor(year) + s(dayInSeason, k=27, by=factor(year))
                  + temperatureResidual + rain + s(windSpeed, k=27) 
                 , data=ragweed
                 , family=quasipoisson
                   )

```

```{r}
fit3 <- mgcv::gam(sqrt(pollenCount) ~ factor(year) + s(dayInSeason, k=27, by=factor(year))
                  + temperatureResidual + rain + s(windSpeed, k=27) 
                 , data=ragweed
                 , family=gaussian(link="identity")
                   )

```

## 10D
Obtain residual plots, analogous to those obtained for fit1 in Part B

```{r}
mgcv::gam.check(fit2)
qqnorm(residuals(fit2, type="scaled.pearson"))
abline(0, 1, col="red")
```

```{r}
mgcv::gam.check(fit3)
qqnorm(residuals(fit3, type="scaled.pearson"))
abline(0, 1, col="red")
```

Based on the above summaries, I believe that the Gaussian fit (fit3) is superior to the Poisson and QuasiPoisson fits.  

```{r}
# graphical representations are available above
mgcv::summary.gam(fit3)
```

## 10F
Come back later

# Question 2

For the square root transformation (fit3), see whether the idea of constant variance is reasonable.
```{r}
resid <- abs(residuals(fit3))
fitted <- fitted(fit3)
plot(fitted, resid
     , main="Fitted Values vs. Absolute Residuals - Fit 3"
     , xlab="Fitted Values"
     , ylab="Abs. Residuals"
     )

```

From the above plot, it does not appear that the assumption of constant variance is reasonable. 

# Question 3
From the estimated scale in the QuasiPoisson model, does the scale parameter imply that we may have concern about overdispersion?

```{r}
fit2$scale
```

Considering that we are looking for a scale value of 1, I believe we do have legitimate concerns about overdispersion.

## Question 4
Test whether the QuasiPoisson fit is statistically significantly different from the Poisson fit.

```{r}
anova(fit1, fit2, test="Chisq")
```

It appears that we have statistically significant differences between the Poisson and QuasiPoisson fits.

# Question 5
In the Poisson fit, is there evidence that there is a year by dayInSeason interaction?

```{r}
mgcv::summary.gam(fit1)
```

```{r}
smaller_fit1 <- mgcv::gam(pollenCount ~ factor(year) + s(dayInSeason, k=27)
                  + temperatureResidual + rain + s(windSpeed, k=27) 
                 , data=ragweed
                 , family=poisson
                   )
anova(fit1, smaller_fit1, test="Chisq")

```

Both evaluations suggest significant evidence of a year by dayInSeason interaction.

# Question 6
Remove the interaction from the model, still doing the quasipoisson; do a stepwise regression - which model is seleted?

```{r}
detach("package:mgcv", unload=TRUE)
library("gam")
```

```{r}
str(ragweed)
```

```{r}
quasi_init <- gam::gam(pollenCount ~ factor(year) + dayInSeason
                  + temperatureResidual + rain + windSpeed 
                 , data=ragweed
                 , family=quasipoisson(link="log")
                   )


quasi_step <- gam::step.gam(quasi_init, scope=list(
                        "year" = ~1 + year
                        , "dayInSeason" = ~1 + dayInSeason + s(dayInSeason, 2)
                        , "temperatureResidual" = ~1 + temperatureResidual +  s(temperatureResidual, 2)
                        , "rain" = ~1 + rain + s(rain, 2)
                        , "windspeed" = ~1 + windspeed + s(windspeed, 2)
                        )
                        , 
                          )



```

# Question 7

```{r}
str(ragweed)
```

```{r}
Y <- ragweed$pollenCount
Y_alt <- log(ragweed$pollenCount)
X <- ragweed[, c(2, 3, 5, 6, 7)]
```

```{r}
cosso_mod <- cosso::cosso(x=X, y=Y_alt, family="Gaussian")
```

