---
title: "Assignment 15; STAT 689"
author: "Philip Anderson; panders2@tamu.edu"
date: "4/21/2018"
output: pdf_document
---

```{r setup, include=FALSE}
library("tidyverse")
library("mgcv")
```

```{r}
rm(list=ls())
```

```{r}
# bring in data
sim <- read.csv("/Users/panders2/Documents/schools/tamu/stat_689/homework/semiparametric-regression/misc/OPEN_sim.csv")
names(sim) <- tolower(names(sim))
str(sim)
```

# Question 1
Run a random-intercept logistic spline regression with Y=indicator that Bio < 27.5, X=FFQ (spline), and Z=(Age, BMI) (linear).

```{r}
# generate a binary class for Biomarkers
sim$bio_bin <- ifelse(sim$bio < 27.5, 1, 0)
# check
summary(sim[sim$bio_bin==1, ]$bio)
summary(sim[sim$bio_bin==0, ]$bio)
```

```{r}
mod_one <- mgcv::gamm(bio_bin ~ s(ffq) + age + bmi
                      , random=list(id = ~ 1)
                      , family=binomial
                      , data=sim
                      )
```

# Question 2
Which among X and Z are statistically significant predictors?

```{r}
summary(mod_one$gam)
```

Of the included predictor variables, only _X_, or the smoothed _FFQ_ term is statistically significant.

# Question 3
Graph the fitted probabilities for people who are 55 years old and whose BMI = 25.

```{r}
ng <- 101
cfpg <- seq(from=min(sim$ffq)
            , to=max(sim$ffq)
            , length=ng
              )
newData <- as.data.frame(cbind(cfpg
              , rep(55, ng)
              , rep(25, ng)
                              )
                        )
names(newData) <- c("ffq", "age", "bmi")
newDataList <- as.list(newData)

predObj <- predict(mod_one$gam, newdata=newDataList, se=T)

muHatg <- 1/(1+exp(-predObj$fit))
aa     <- predObj$fit + 2*predObj$se
bb     <- predObj$fit - 2*predObj$se
lowergg <- 1 / (1 + exp(-bb)) 
uppergg <- 1 / (1 + exp(-aa)) 

```

```{r}
plot(cfpg, muHatg, type="n", main="Fitted Probabilities for Individuals 55 years old with BMI=25")
polygon(c(cfpg, rev(cfpg))
        , c(lowergg, rev(uppergg))
        , col="cadetblue1"
        , border=F
        )
lines(cfpg, muHatg, lwd=2)
rug(sim$ffq)
```




