---
title: "Assignment_06 - STAT 689"
author: "Philip Anderson; panders2@tamu.edu"
date: "2/14/2018"
output: pdf_document
---

```{r, message=F}
# import third-party modules
library("HRW")
library("tidyverse")
library("mgcv")
library("nlme")
```

Read in our data
```{r}
fram <- read.csv("/Users/panders2/Documents/schools/tamu/stat_689/homework/semiparametric-regression/hw06/framington.csv")
names(fram) <- tolower(names(fram))
# check that we have 1615 observations
dim(fram)

```

Create new variables for the systolic blood pressure readings and the two cholesterol measurements.

```{r}
# systolic blood pressure first
fram$lsbp <- log(((fram$sbp21 + fram$sbp22 + fram$sbp31 + fram$sbp32) / 4) - 50)
# cholesterol measurements second
fram$lcholest <- log((fram$cholest2 + fram$cholest3) / 2)
``` 

Im guessing that we conducted the transformations to take the variables to normality.  Checking that here.
```{r}
plot(density(fram$lsbp), main="Blood Pressure Density Plot")
plot(density(fram$lcholest), main="Cholesterol Density Plot")
```

Keep only the variables that we will be working with directly and make sure everything seems reasonable.

```{r}
fram2 <- fram %>%
  dplyr::select(chd, age, smoker, lsbp, lcholest)
head(fram2)
```


# Question 1
Fit a multiple linear regression of LSBP on lcholest and smoker via "lm" function.  Produce estimates, standard errors, and p-values.

```{r}
# fit model
lin_mod <- lm(lsbp ~ smoker + lcholest
              , data=fram2
              )
# produce summary
summary(lin_mod)
```


# Question 2
Conduct web research on whether smokers have higher or lower blood pressure on average compared with non-smokers.

WebMD indicates that individuals who smoke tend to have higher blood pressure than those who do not.  This is not consistent with my findings from Question 1, which indicate that participants who smoke have lower blood pressure than those who do not, conditional on our transformed cholesterol variable.  There is nothing in the documentation to indicate that smoker is not encoded with '1' as the positive class.  This result is suspicious, and suggests that we need to check our data or revisit our model specification. 

# Question 3
The model produces the expectation of LSBP given smoking status _conditional on_ cholesterol.

# Question 4
Recreate the same model as in Question 1, but also include an interaction amongst the independent variables.

```{r}
lin_mod2 <- lm(lsbp ~ smoker + lcholest + smoker:lcholest
              , data=fram2
              )

summary(lin_mod2)
```

The smoking indicator is still negatively associated with our blood pressure variable, but is no longer significant in the presence of the interaction term.

# Question 5
Run a semiparametric ANCOVA with mgcv, the semiparametric version of an ANCOVA without an interaction.

```{r}
semi_mod <- mgcv::gam(lsbp ~ factor(smoker) +
                        s(lcholest, k=23, bs="cr")
                      , data=fram2
                      , method="REML"
                     )
summary(semi_mod)
```

# Question 6
For the data in part 5, display a plot of the two lines, but without the data

```{r} 
# first, generate the data required for plotting
ng <- 1000
# x var
x_var <- seq(from=min(fram2$lcholest)
             , to=max(fram2$lcholest)
             , len=ng
             )
# y prediction variables
f_hat_smoker <- predict(semi_mod
                        , newdata=data.frame(
                         smoker=rep('1', ng)
                         , lcholest=x_var
                        )
                        , se.fit=TRUE
                        )

f_hat_nosmoke <- predict(semi_mod
                         , newdata=data.frame(
                            smoker=rep('0', ng)
                            , lcholest=x_var
                           ) 
                         , se.fit=TRUE
                         )
```

```{r}
lineColors <- c("darkolivegreen", "dodgerblue")

plot(fram2$lcholest, fram2$lsbp, type="n"
     , xlab="Cholesterol Metric"
     , ylab = "Blood Pressure Metric"
     , main="Blood Presure by Cholesterol ANCOVA comparison"
       )

lines(x_var, f_hat_smoker$fit, lty=2, lwd=2.5, col=lineColors[1])
lines(x_var, f_hat_nosmoke$fit, lty=3, lwd=2.5, col=lineColors[2])

legend(4.8, 5
       , c("Smoker", "NonSmoker")
       , lty=c(2,3)
       , lwd=rep(2.5, 2)
       , col=c(lineColors[1], lineColors[2])
       )
```

# Question 7
Display a plot of the two lines, but split into two separate graphics, and with pointwise 95% confidence intervals.

```{r}
plot_with_bands <- function(mod_pred, range_var, color_vec, color_idx, lty, title="NULL") {
  # create a regression fit with 95% pointwise confidence intervals for the line
  upper <- mod_pred$fit + (1.96 * mod_pred$se.fit)
  lower <- mod_pred$fit - (1.96 * mod_pred$se.fit)
  fit <- mod_pred$fit
  
  plot(fram2$lcholest, fram2$lsbp, type="n"
       , xlab="Cholesterol Metric"
       , ylab = "Blood Pressure Metric"
       , main=title
         )

  lines(range_var, fit, lty=2, lwd=2.5, col=color_vec[color_idx])

  polygon(x=c(range_var,rev(range_var))
        , y=c(upper
        , rev(lower))
        , col="gray"
        , border=NA
        , density=75
        )

                  }
# execute the function 
par(mfrow=c(1,2))
plot_with_bands(mod_pred=f_hat_smoker, range_var=x_var, color_vec=lineColors, color_idx=1, lty=2
                , title="Smoker")
plot_with_bands(mod_pred=f_hat_nosmoke, range_var=x_var, color_vec=lineColors, color_idx=2, lty=3
                , title="Non-Smoker")
par(mfrow=c(1,1)) # reset

```


# Question 8
Run the semiparametric version of ANCOVA but with an interaction.  

```{r}
semi_mod2 <- mgcv::gam(lsbp ~ factor(smoker)*lcholest +
                        s(lcholest, k=23, bs="cr")
                      , data=fram2
#                      , method="REML"
                     )
summary(semi_mod2)

```

From the above results, it does not appear that there is any sort of interaction present between _smoker_ and _lcholest_.  The estimate for this term is negligibly small, and the p-value is very large, indicating a lack of significance. 


# Question 9
Display the fits of the above regressions, but without the data points.

```{r}
ng <- 1000
x_vec <- seq(from=min(fram2$lcholest), to=max(fram2$lcholest), len=ng)
fHat_smoke <- predict(semi_mod2, newdata=data.frame(
                       lcholest=x_vec
                       , smoker=rep('1', ng)
                        )
                      )

fHat_nosmoke <- predict(semi_mod2, newdata=data.frame(
                       lcholest=x_vec
                       , smoker=rep('0', ng)
                        )
                      )
```

```{r}
plot(fram2$lcholest, fram2$lsbp, type='n'
     , xlab="Cholesterol"
     , ylab="Blood Pressure"
     , main="Cholesterol by Blood Pressure"
     )
mtext("Semi-parametric ANCOVA with Interaction")

col_vec <- c("dodgerblue", "forestgreen")
lines(x_vec, fHat_smoke, col=col_vec[1], lwd=2.5, lty=2)
lines(x_vec, fHat_nosmoke, col=col_vec[2], lwd=2.5, lty=3)

legend(4.8, 5.2, c("Smoke", "NoSmoke")
       , col=col_vec
       , lwd=rep(2.5, 2)
       , lty=c(2,3)
      )
``` 


# Question 10
What does the interaction mean in the case when the factors are binary?

When we have a binary factor for our ANCOVA model, this indicates that the interaction term's coefficient is reflecting what happens to our outcome variable for that factor's non-reference class only.  In this case, we are seeing the results of _smoker_==1 interacting with _lcholest_.

# Question 11
Run an analysis of whether our two regression lines are significantly different.

```{r}
# First, fit the null model - this is run on both the smokers and non-smokers
contrast_mod1 <- mgcv::gam(lsbp ~ s(lcholest), data=fram2)

# indicator of the smoke varible taking positive class
smoke_ind <- as.numeric(fram2$smoker==1)
# now, fit the alternative model
contrast_mod2 <- mgcv::gam(lsbp ~ s(lcholest, smoke_ind), data=fram2)

# run the final test
anova(contrast_mod1, contrast_mod2, test="F")


```

It appears that there is a significant difference between the two fits, for smoker and non-smoker.

# Question 12
Ignore the smokers, and run native code using lme to fit LSBP vs. age.  

```{r}
fram_ns <- fram2[which(fram2$smoker==0), ]
x <- fram_ns$age
y <- fram_ns$lsbp

numIntKnots <- 23
intKnots <- quantile(unique(x), seq(0, 1, length=numIntKnots+2))[-c(1, numIntKnots+2)]

a <- 1.01 * min(x) - 0.01*max(x)
b <- 1.01 * max(x) - 0.01*min(x)
Z <- HRW::ZOSull(x, range.x=c(a, b), intKnots=intKnots)

dummyID <- factor(rep(1, length(x)))

# prep is finally over - run the model
mm_fit <- nlme::lme(y ~ x, random=list(dummyID=pdIdent(~ -1 + Z)))

```

# Question 13
Display the fit without data points.  Also provide a confidence band.

```{r}
# extract our model coefficients
betaHat <- mm_fit$coefficients$fixed
uHat <- unlist(mm_fit$coefficients$random)
# create plotting parameters
ng <- 1001
xg <- seq(a, b, length.out=ng)
Xg <- cbind(rep(1, ng), xg)
Zg <- HRW::ZOSull(xg, range.x=c(a,b), intKnots=intKnots)
fHatg <- as.vector(Xg %*% betaHat + Zg %*% uHat)
plot(x, y, type='n'
     , xlab="Age"
     , ylab="Blood Pressure"
     , main='Blood Pressure by Age Mixed Model Fit'
     )
# plot the main line
lines(xg, fHatg, col="darkgreen", lwd=2.5)

# grab the standard error estimates
se_pred <- diag(Xg %*% mm_fit$varFix %*% t(Xg))
SE <- sqrt(se_pred)
# lower and upper bounds
upper <- fHatg + (1.96 * se_pred)
lower <- fHatg - (1.96 * se_pred)
# plot the confidence polygons
polygon(x=c(xg,rev(xg))
      , y=c(upper
      , rev(lower))
      , col="gray"
      , border=NA
      , density=75
      )

```

Pointwise 95% confidence interval is there, but very narrow and difficult to see.

# Question 14
Is the fit of our mixed model statistically significant?

```{r}
anova(mm_fit)
```

The F-test indicates that the fit of our model is significant.

# Question 15
Test whether the fit is linear or quadratic vs. the need to do a semiparametric fit.

```{r}
mm_fit <- nlme::lme(y ~ x, random=list(dummyID=pdIdent(~ -1 + Z)))
mm_fit2 <- nlme::lme(y ~ x**2, random=list(dummyID=pdIdent(~ -1+Z)))
anova(mm_fit, mm_fit2)
```


# It does not appear that the quadratic fit is improving our model in any appreciable way.

